BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-01-07 Paper/BOAST.tex
rework debugging section

@@

  \subsubsection{Multi-target Language Generation}

Language availability and performance can vary between platforms. It is thus
important to be able to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages can have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

@@

In order to use this methodology one has to be able to trace an application to
get input data. Such a tracer, dedicated to CUDA and OpenCL, will be presented
in the SPECFEM3D use case.

@@

For these reasons, a convolution library has been engineered with the help of
BOAST.  \textbf{The detailed API of the library will be described elsewhere}.  The
spirit is similar to BLAS-LAPACK API, where low-level operations are scheduled
and called from high-level operations.  Each of the building blocks of these
convolution libraries is optimally tuned by BOAST by choosing the sources
providing the optimal kernel for the chosen computing platform.  The sources of
these kernels are then collected and compiled to meet the API of the library.

>>>>

\newcommand{\productname}[1]{\textsc{#1}}
\newcommand{\latin}[1]{\textit{#1}}

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\eg}{\latin{e.g.,}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

\newcommand{\etc}[1]{\latin{etc}}

@@

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance can vary between platforms. It is thus
important to be able to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages can have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

@@

In order to use this methodology one has to be able to trace an application to
get input data. Such a tracer, dedicated to CUDA and OpenCL, will be presented
in the next subsection.

\subsection{Trace Debugging of BOAST-based GPU Applications}
\label{sec:gputrace}

Debugging applications running on GPU environments is well-recognized
as a hard and time-consuming activity. In complement with BOAST, we
designed a trace-based debugging tool that simplifies this porting
operation. The tool relies on BOAST support of multi-target code
generation (Section~\ref{sec:multitarget}), used to port an
application from one GPU-programming framework to another. A
case-study of the portage of a Cuda application to OpenCL is presented
in Section~\ref{subsec:specfem}.

The idea behind the tool is based on the assumption that the different
GPU ports of the code should do the very same operations, at least at
the logical the logical level (the APIs will have
\emph{implementation} differences, but they should offer nonetheless
same functionalities). The usage of BOAST framework sustains this
assumption, as both sets of kernels should generated from the same
source code.

Hence, the verificatation and validation of the new port can be
narrowed down to asserting that both codes apply the same operations
on the GPU. And the debugging part will consist in understanding what
diverges. To that purpose, we developed \code{GPUTrace}, inspired by
\code{strace} and \code{ltrace} tools: \code{GPUTrace} dynamically
preloads a library between the application and the GPU library, and
collects the function name, execution range and argument values (input
\emph{and} output) of the relevant function calls. These information
are traced in a unified format for all the APIs.

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of API
parameters are handles to opaque types, so, in order to provide
meaningful information, \code{GPUTrace} stores information about these
objects (handle value, buffer creation size and attributes, kernel
name and prototype, \etc{}).

Once two call traces have been generated by \code{GPUTrace}, it is
easy to compare them with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}.

By default, \code{GPUTrace} only prints a unique identifier (the
creation index) for the memory buffers. A first flag can be activated
to append the first bits of the buffer to the trace, printed in the
right format, for visual inspection. Or another flag can be set to
dump the whole content of the buffer into a file, for a full
inspection.

This last option is also useful for generating \emph{replay buffers}
for BOAST kernel execution. With a set of filters based on kernel
names and execution counters, developers can precisely select which
kernel execution parameters should be dumped, for further reuse as
real-case benchmarks and non-regression testing.

@@

For these reasons, a convolution library has been engineered with the help of
BOAST.  The detailed API of the library will be described elsewhere.  The
spirit is similar to BLAS-LAPACK API, where low-level operations are scheduled
and called from high-level operations.  Each of the building blocks of these
convolution libraries is optimally tuned by BOAST by choosing the sources
providing the optimal kernel for the chosen computing platform.  The sources of
these kernels are then collected and compiled to meet the API of the library.

### 2015-01-07 Paper/specfem.tex
work on specfem/gputrace section
rework debugging section

\newcommand{\productname}[1]{\textsc{#1}}
\newcommand{\latin}[1]{\textit{#1}}

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\eg}{\latin{e.g.,}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

\newcommand{\etc}[1]{\latin{etc}}

\subsection{Porting SPECFEM3D to OpenCL using BOAST}

@@

As part of the Mont-Blanc project, we had to port \Specfem to \OCL, so
that it could be use to benchmark Mont-Blanc HPC platforms.

@@

\subparagraph{Porting kernels to BOAST} Nvidia \Cuda and \OCL are
based on the same programming model: a massively parallel accelerator
running in another memory environment. Thanks to that proximity, we,
BOAST developers, have been able to carry out most of the porting task
with only a limited knowledge about \Specfem internal physics.

That lack of knowledge led us to be particularly attentive to the path
we undertook for the porting, as we would have been unable to
understand how and why the application was not operating properly, if
it was to fail.

@@

Hence, our idea to locate the execution problems was to make sure that
both execution actually did the same thing. As the \OCL results were
invalid, we knew the executions would diverge at one or several
points.

\subparagraph{Debugging \OCL Execution: \code{GPUTrace}} In order to
ensure that \Cuda and \OCL executions were doing the same operations,
we developed a tool that could intercept all the function calls
related to the GPU and store them for offline processing, in a format
common to \OCL and \Cuda. Our solution was inspired by \code{strace}
and \code{ltrace} tools: it dynamically preloads a library (with the
help \code{LD\_PRELOAD} in \productname{GNU/Linux}-based environments)
between the application and the GPU library, and prints the arguments
(input \emph{and} output) of the relevant function calls.

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of \OCL and
\Cuda parameters are handles to opaque types, so, in order to provide
more meaningful information, \code{GPUTrace} stores internally
information about these handles (buffer creation size and index,
kernel name and prototype, \etc{}). % more details here?

Once \code{GPUTrace} was operational, it was easy to confront \Cuda
and \OCL outputs with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}. One last problem remained, impossible to
miss because of the seismograms not matching (but with just a slight
drift). We added more verbosity to \code{GPUTrace} output: first the
initial bits of the GPU memory buffers, then their full content. The
drift was visible in the trace, but it was nonetheless unclear where
it started. We finally got it after hours of code review of BOAST
kernels and \OCL code: one kernel was \emph{three}-dimensional,
whereas the others were two-dimensional. But for all of them, only two
dimensions were passed, and one was missing. As it was the same person
who coded the \OCL port and the GPU tracer, this third dimension was
forgotten in both codes\ldots

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and x Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL runtime, as of
version 1.1.

>>>>

\subsection{Porting SPECFEM3D to OpenCL using BOAST}
\label{subsec:specfem}

@@

As part of the Mont-Blanc project, we had to port \Specfem to \OCL, so
that it could be use to benchmark Mont-Blanc HPC platform.

@@

\subparagraph{Porting kernels to BOAST}

Nvidia \Cuda and \OCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we, BOAST developers,
have been able to carry out most of the porting task with only a
limited knowledge about \Specfem internal physics.

This lack of \Specfem internal knowledge led us to be particularly
attentive to the path we undertook for the porting, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

@@

Hence, our idea for locating the execution problems was to make sure
that both execution actually did the same thing. As the \OCL results
were invalid, we knew the executions would diverge at one or several
points.

\subparagraph{Debugging \OCL Execution: \code{GPUTrace}}

With the help of \code{GPUTrace} (Section~\ref{sec:gputrace}), we
could to confront \Cuda and \OCL execution traces with a graphical
\code{diff} tool, and spot the different porting mistakes: some
parameters reversed, offsets incorrectly applied, \etc{}.

One last problem remained, clearly highlighted by the seismograms not
matching perfectly (they had a similar shape, but with a reduced
intensity). We added more verbosity to \code{GPUTrace} output: first
the initial bits of the GPU memory buffers, then their full
content. The drift was visible in the trace, but it was nonetheless
unclear where it started. We finally got it after hours of code review
of BOAST kernels and \OCL code: one kernel was
\emph{three}-dimensional, whereas the others were two-dimensional. But
for all of them, only two dimensions were passed, and one was
missing. As it was the same person who coded the \OCL port and the GPU
tracer, this third dimension was forgotten in both codes\ldots

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL runtime, as of
version 1.1.
